### 2020-03-04 rebuttal/rebuttal.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% END MACROS SECTION%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\section*{Response to Reviewer for PONE-D-19-32846 ``Theoretical Investigation of a Genetic Switch for Metabolic Adaptation''}
Kathrin S. Laxhuber,$^{1}$ Muir J. Morrison,$^{2}$ Griffin Chure,$^{3}$ Rob Phillips$^{2, 3, \dagger}$
\vspace{10pt}

\noindent\tiny{$^1$ Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland,
$^2$ Department of Physics, California Institute of Technology, Pasadena, CA, USA,
$^3$ Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA, $^\dagger$ E-mail: \texttt{phillips@pboc.caltech.edu}}

{\Large\textcolor{red}{ToDo: change author list above and at the very end of this.}}

@@

\begin{review}
Second, while the authors do technically recapitulate the bistable
behavior of the system in the stochastic version, the bistability seems
based on their description to be quite fragile and not very meaningful,
since it only occurs here for small distances between the lower and
upper fixed points
\end{review}

\begin{response}
We thank the reviewer for bringing this point to our attention, which we
believe to be a misunderstanding that resulted from a poor explanation
in our original text. We believe we have clarified this in the
manuscript, and allow us elaborate here too: Stochastic bimodality occurs
only when two of the three fixed points are close together, or in other
words, when the system is poised ``sufficiently near'' one of the
bifurcation points. This may become clearer when looking at the
bifurcation diagram that we added to our section about hysteresis at the
end of the main text. For convenience, we insert this plot below in
Fig~\ref{bifucation} which addresses what we believe the reviewer is
referring to: only when two of the three fixed points are close do the
stochastic trajectories occupy both fixed points, resulting in
bimodality.

Since this bimodality only occurs for a small region in parameter space,
this actually indicates strong and not fragile bistability. In other
words, the fact that the system does not stochastically ``jump'' to the
other fixed point means that the fixed points are very stable to
stochastic fluctuations. Thus, the switch-like feature is robust:
there are two distinct states that the system can occupy, and only near
the bifurcation points is it likely for stochastic trajectories to
transition between them. Put differently, deterministic bistability does
not need to result in stochastic bimodality, and in fact stochastic
bimodality is contrary to robust bistability.
\end{response}

\begin{figure}[h!]
	\centering
	\includegraphics{Fig10_bifurcation.eps}
	\vspace*{0.5em}
	\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
		The blue line shows the positions of the deterministic fixed points, while the dashed line indicates the instability of the middle one. The orange and yellow points show the mean of the stochastic simulations when started at zero and at the high fixed point, respectively. The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines. To make this clearer, the bimodal distributions themselves are shown on the right. The arrows illustrate the hysteresis loop in the stochastic simulations.}
	\label{bifucation}
\end{figure}

@@

\begin{response}
We hope that we have addressed all these points in our responses above
and shortly summarize here: We put in some cautionary words about the
general usage of continuous models (beginning of sections on deterministic as well as stochastic results) and demonstrated that for our system
and the conclusions we draw, the continuous model is surprisingly appropriate (bifurcation diagram and time evolution, end of results section).
Furthermore, we extended our characterization of the stochastic model as
suggested by investigating the distribution of adaptation/switching
times (end of results section) as well as by incorporating transcriptional bursting in our
simulation (end of section C in S1 Text). The deterministic and stochastic model are now directly
compared (for the same set of parameters) in the bifurcation diagram as
well as in terms of their adaptation/switching times.
\end{response}

\begin{review}
My other major difficulty here is that essentially all of the work is
premised on reference 22, which is 'unpublished data.' This is the only
experimental evidence given for the bistability of xapAB expression,
justification of many of the model parameters, experimentally expected
switching timescales, etc. It is impossible to properly interpret the
present work, either for the purposes of review or from the viewpoint of
an interested reader, without availability of those data, which either
ought to be their own independent publication or ought to be included
with this one.
\end{review}

\begin{response}
With further consideration, we agree with the reviewer that we relied
rather heavily on this unpublished data and that it is therefore
essential to make the experimental data available.
We have opted to include the data with this publication rather than
separately. For this reason, we have added Griffin Chure, Nathan Belliveau, Charlotte Strandkvist, and Kyle Naughton to the author list, who collected or were involved in collecting the data. We have added a new section, ``experimental motivation,'' in
the beginning of the main text which discusses the main features of the
data that are relevant for our model. Furthermore, we added a section on
the materials and experimental methods at the end of S1 Text (section
E).
The raw data is
publicly available at DOI: 10.5281/zenodo.3695049.
\end{response}

### 2020-03-05 rebuttal/rebuttal.tex

\mdfdefinestyle{respstyle}{
outerlinewidth=2pt,
bottomline=false,
leftline=false,rightline=false,
topline=false,
font=\small,
backgroundcolor=gray!20,
skipabove=\baselineskip,
skipbelow=\baselineskip,
roundcorner=10pt,
innertopmargin=0.5em,
innerbottommargin=1em,
splittopskip=1em,
splitbottomskip=1em
}

\newmdenv[style=revstyle,frametitle=\sffamily \itshape Reviewer Comment]{review}
\newmdenv[style=respstyle,frametitle=\sffamily \itshape Author Response]{response}
\newmdenv[style=respstyle,frametitle=\sffamily \itshape Author Response: Executive Summary]{responseSummary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% END MACROS SECTION%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\section*{Response to Reviewer for PONE-D-19-32846 ``Theoretical Investigation of a Genetic Switch for Metabolic Adaptation''}
Kathrin S. Laxhuber\textsuperscript{1},
Muir J. Morrison\textsuperscript{2},
Griffin Chure\textsuperscript{3},
Nathan M. Belliveau\textsuperscript{4},
Charlotte Strandkvist\textsuperscript{5},
Kyle L. Naughton\textsuperscript{6},
Rob Phillips\textsuperscript{2,3$\dagger$}
\vspace{10pt}

\noindent\tiny{
$^1$ Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland,
$^2$ Department of Physics, California Institute of Technology, Pasadena, CA, USA,
$^3$ Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA,
$^4$ Howard Hughes Medical Institute, University of Washington, Seattle, WA, USA,
$^5$ Department of Systems Biology, Harvard Medical School, Boston, MA, USA,
$^6$ Department of Physics and Astronomy, University of Southern California, Los Angeles, CA, USA, $^\dagger$ E-mail: \texttt{phillips@pboc.caltech.edu}}

### 2020-03-05 s1_text/s1_text.tex
Edits to expt details + author list & minor polish

\begin{flushleft}
	{\Large
		\textbf\newline{Supporting information for ``theoretical investigation of a genetic switch for metabolic adaptation''}
	}
	\newline
	% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
		\\
	Kathrin S. Laxhuber\textsuperscript{1},
	Muir J. Morrison\textsuperscript{2},
	Griffin Chure\textsuperscript{3},
	Rob Phillips\textsuperscript{2,3$\ast$}
	\\
	\bigskip
	\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
	\\
	\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
	\\
	\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
	\\
	\bigskip

@@

\section{Experimental materials and methods} \label{sec:exp}
\paragraph{Construction of $\bm{\Delta{xapABR}}$ deletion strain.} Cells containing
plasmid pSIM6 were diluted 1:100 in $50 \unit{mL}$ LB media and grown to an OD600
of $\approx 0.4$ at $30^\circ$C. The culture was immediately placed in
a water bath shaker at $43\unit{^\circ C}$ for 15 minutes and then cooled in an
ice bath for 10 minutes. Cells were then spun down for 10 minutes
($4,000 \unit{g}$, $4\unit{^\circ C}$) and resuspended on ice in $50\unit{mL}$ of chilled
water. This was repeated three times before resuspending in $200 \unit{\upmu L}$
of chilled water to generate competent cells. Homologous primer
extension sequences were obtained from Baba {\it et al.} \cite{Baba2006}
and used to generate linear DNA containing a kanamycin resistance
gene insert by PCR (amplified from plasmid pKD4), using primers that.
contained homology for the region on the chromosome to be deleted
\cite{Datsenko2000}. Electroporation was performed using $1 \unit{\upmu L}$
purified PCR product (about $100 \unit{ng}$ DNA), mixed with $50 \unit{\upmu L}$ cells.
Cells were resuspended in $750 \unit{\upmu L}$ SOC media and outgrowth was
performed on a shaker at $30\unit{^\circ C}$ for 90-120 minutes. Cells were then
selected on an LB-agar plate containing kanamycin ($30 \unit{\upmu g/mL}$) and
grown overnight at $30\unit{^\circ C}$. The deletions were confirmed by both
colony PCR and sequencing. After confirmation, the deletion was
transferred to a clean strain through P1 transduction and selection on
kanamycin. Removal of kanamycin resistance was achieved by transforming
cells with plasmid pCP20 (expressing FLP recombinase, carbenicillin
resistance, on a temperature-sensitive replicon), with cells grown at
$30\unit{^\circ C}$. Lastly, several colonies were then growth colony-purified
once non-selectively at $43\unit{^\circ C}$ and then tested for loss of
antibiotic resistance for both kanamycin and carbenicillin.

@@

\paragraph{Primers for deletion strains}
Generation of $\Delta{xapABR}$:
actatttccggcaattcctgtctcctcacctactgtgtcaatgcagccaaCATATGAATATCCTCCTTAG; $xapABR<>P2$
(lower case is homology, upper case is pKD4 primer)

agatacaacaggtatcgctaatatagaggaaagtgtcagagttaaaagatgttttttcatGTGTAGGCTGGAGCTGCTTC; $xapABR<>P1$ (lower case is homology, upper case is pKD4 primer)

Primers for colony-PCR to confirm deletion of xapABR:
(upstream of xapR) cgggtcgttagctcagttggta
(downstream of XapA): ttcccagtcgaagaacccgtac

\textbf{Finish reformatting to table, add other primers?}

@@

	\bibitem{Datsenko2000}
	Datsenko, KA, Wanner, BL.
	\newblock One-step inactivation of chromosomal genes
	in Escherichia coli K-12 using PCR products.
	\newblock PNAS, 2000;97(12), 6640–6645.
	\newblock doi:{10.1073/pnas.120163297}

>>>>

\begin{flushleft}
	{\Large
		\textbf\newline{Supporting information for ``theoretical investigation of a genetic switch for metabolic adaptation''}
	}
	\newline
	% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
		\\
	Kathrin S. Laxhuber\textsuperscript{1},
	Muir J. Morrison\textsuperscript{2},
	Griffin Chure\textsuperscript{3},
	Nathan M. Belliveau\textsuperscript{4},
	Charlotte Strandkvist\textsuperscript{5},
	Kyle L. Naughton\textsuperscript{6},
	Rob Phillips\textsuperscript{2,3$\ast$}
	\\
	\bigskip
	\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
	\\
	\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
	\\
	\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
	\\
	\textbf{4} Howard Hughes Medical Institute, University of Washington, Seattle, WA, USA
	\\
	\textbf{5} Department of Systems Biology, Harvard Medical School, Boston, MA, USA
	\\
	\textbf{6} Department of Physics and Astronomy, University of Southern California, Los Angeles, CA, USA
	\\
	\bigskip

@@

\section{Experimental materials and methods} \label{sec:exp}
\paragraph{Construction of $\bm{\Delta{xapABR}}$ deletion strain.} Cells containing
plasmid pSIM6 were diluted 1:100 in $50 \unit{mL}$ LB media and grown to an OD600
of $\approx 0.4$ at $30^\circ$C. The culture was immediately placed in
a water bath shaker at $43\unit{^\circ C}$ for 15 minutes and then cooled in an
ice bath for 10 minutes. Cells were then spun down for 10 minutes
($4,000 \unit{g}$, $4\unit{^\circ C}$) and resuspended on ice in $50\unit{mL}$ of chilled
water. This was repeated three times before resuspending in $200 \unit{\upmu L}$
of chilled water to generate competent cells. Homologous primer
extension sequences were obtained from Baba {\it et al.} \cite{Baba2006}
and used to generate linear DNA containing a kanamycin resistance
gene insert by PCR (amplified from plasmid pKD4), using primers that
contained homology for the region on the chromosome to be deleted
\cite{Datsenko2000}. Electroporation was performed using $1 \unit{\upmu L}$
purified PCR product (about $100 \unit{ng}$ DNA), mixed with $50 \unit{\upmu L}$ cells.
Cells were resuspended in $750 \unit{\upmu L}$ SOC media and outgrowth was
performed on a shaker at $30\unit{^\circ C}$ for 90-120 minutes. Cells were then
selected on an LB-agar plate containing kanamycin ($30 \unit{\upmu g/mL}$) and
grown overnight at $30\unit{^\circ C}$. The deletions were confirmed by both
colony PCR and sequencing. After confirmation, the deletion was
transferred to a clean strain through P1 transduction and selection on
kanamycin. Removal of kanamycin resistance was achieved by transforming
cells with plasmid pCP20 (expressing FLP recombinase, carbenicillin
resistance, on a temperature-sensitive replicon), with cells grown at
$30\unit{^\circ C}$. Lastly, several colonies were then growth colony-purified
once non-selectively at $43\unit{^\circ C}$ and then tested for loss of
antibiotic resistance for both kanamycin and carbenicillin.

@@

\begin{table}
\begin{tabular}{c p{5cm} p{4cm}}
Primer & Sequence & Comments \\
\hline
$xapABR<>P2$ &
actatttccggcaattcctgtctcctcac ctactgtgtcaatgcagccaaCATAT GAATATCCTCCTTAG &
Forward primer for pKD4 with homology to \textit{xapABR} locus
\\
$xapABR<>P1$ &
agatacaacaggtatcgctaatatagagg aaagtgtcagagttaaaagatgtttt ttcatGTGTAGGCTGGAGCTGCTTC &
Reverse primer for pKD4 with homology to \textit{xapABR} locus
\\
xapABR-seq-fwd &
cgggtcgttagctcagttggta &
upstream of \textit{xapR}, to check \textit{xapABR} deletion
\\
xapABR-seq-rev &
ttcccagtcgaagaacccgtac &
downstream of \textit{xapA}, to check \textit{xapABR} deletion
\end{tabular}
\caption{\textbf{Primers for deletion strains.}
Lower case sequences are homologous to a chromosomal locus.
Upper case indicates sequences used for PCR amplification.
}
\end{table}

@@

	\bibitem{Datsenko2000}
	Datsenko KA, Wanner BL.
	\newblock One-step inactivation of chromosomal genes
	in Escherichia coli K-12 using PCR products.
	\newblock PNAS, 2000;97(12), 6640–6645.
	\newblock doi:{10.1073/pnas.120163297}

	\bibitem{Brewster2014}
	Jones, DL, Brewster, RC, Phillips, R
	\newblock Promoter architecture dictates cell-to-cell variability in gene expression.
	\newblock Science, 2014;346(6216), 1533--1536.
	\newblock doi:{10.1126/science.1255301}
